---
title: "outlierMBC"
author: "Ultán P. Doherty"
date: "2024-06-26"
output: github_document
---

# Outlier Identification for Model-Based Clustering

* `ombc_gmm` -  Identify multivariate outliers while clustering data with a Gaussian mixture model.
* `ombc_lcwm` - Identify covariate and/or response outliers while fitting a linear cluster-weighted model to the data.
* `simulate_gmm` - Simulate data from a Gaussian mixture model with multivariate outliers.
* `simulate_lcwm` - Simulate data from a linear cluster-weighted model with covariate and/or response outliers.


```{r, include=FALSE}
library(ggplot2)
devtools::load_all()
```

## Mixture Models

```{r}
gmm_p2g3 <- simulate_gmm(
  n = c(2000, 1000, 1000),
  mu = list(c(-1, 0), c(+1, -1), c(+1, +1)),
  sigma = list(diag(c(0.2, 4 * 0.2)), diag(c(0.2, 0.2)), diag(c(0.2, 0.2))),
  outlier_num = 40,
  seed = 123,
  crit_val = 0.9999,
  unif_range_multiplier = 1.5
)

ombc_gmm_p2g3 <- ombc_gmm(
  gmm_p2g3[, 1:2],
  comp_num = 3,
  max_out = 80
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:80,
  distrib_diffs = ombc_gmm_p2g3$distrib_diffs
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = ombc_gmm_p2g3$outlier_num) +
  labs(x = "Number of Potential Outliers Removed")

gmm_p2g3 |>
  ggplot(aes(
    x = X1, y = X2,
    colour = as.factor(ombc_gmm_p2g3$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1, 2, 3))
```

## Linear Cluster-Weighted Models

### Single Component, Response Outliers

```{r}
lcwm_p1g1_y_only <- simulate_lcwm(
  n = 1000,
  mu = list(c(1)),
  sigma = list(as.matrix(0.1)),
  beta = list(c(1, 1)),
  error_sd = 0.5,
  outlier_num = 20,
  outlier_type = "y_only",
  seed = 123,
  crit_val = 0.9999
)

ombc_lcwm_p1g1_y_only <- ombc_lcwm(
  xy = lcwm_p1g1_y_only,
  x = lcwm_p1g1_y_only$X1,
  y_formula = Y ~ X1,
  comp_num = 1,
  max_out = 40,
  mnames = "V",
  seed = 123,
  outlier_type = "y_only"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:40,
  distrib_diffs = ombc_lcwm_p1g1_y_only$distrib_diffs,
  outlier_num = ombc_lcwm_p1g1_y_only$outlier_num
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs <- 
  ombc_lcwm_p1g1_y_only$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients

lcwm_p1g1_y_only |>
  ggplot(aes(
    x = X1, y = Y,
    colour = as.factor(ombc_lcwm_p1g1_y_only$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  geom_abline(
    aes(intercept = coeffs[1], slope = coeffs[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1))
```

### Single Component, Covariate Outliers

```{r}
lcwm_p1g1_x_only <- simulate_lcwm(
  n = 1000,
  mu = list(c(1)),
  sigma = list(as.matrix(0.1)),
  beta = list(c(1, 1)),
  error_sd = 0.5,
  outlier_num = 20,
  outlier_type = "x_only",
  seed = 123,
  crit_val = 0.9999
)

ombc_lcwm_p1g1_x_only <- ombc_lcwm(
  xy = lcwm_p1g1_x_only,
  x = lcwm_p1g1_x_only$X1,
  y_formula = Y ~ X1,
  comp_num = 1,
  max_out = 40,
  mnames = "V",
  seed = 123,
  outlier_type = "x_only"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:40,
  distrib_diffs = ombc_lcwm_p1g1_x_only$distrib_diffs,
  outlier_num = ombc_lcwm_p1g1_x_only$outlier_num
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs <- 
  ombc_lcwm_p1g1_x_only$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients

lcwm_p1g1_x_only |>
  ggplot(aes(
    x = X1, y = Y, 
    colour = as.factor(ombc_lcwm_p1g1_x_only$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  geom_abline(
    aes(intercept = coeffs[1], slope = coeffs[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1))
```

### Single Component, Combined Outliers

```{r}
lcwm_p1g1_x_and_y <- simulate_lcwm(
  n = 1000,
  mu = list(c(1)),
  sigma = list(as.matrix(0.1)),
  beta = list(c(1, 1)),
  error_sd = 0.5,
  outlier_num = 20,
  outlier_type = "x_and_y",
  seed = 123,
  crit_val = 0.9999
)

ombc_lcwm_p1g1_x_and_y <- ombc_lcwm(
  xy = lcwm_p1g1_x_and_y,
  x = lcwm_p1g1_x_and_y$X1,
  y_formula = Y ~ X1,
  comp_num = 1,
  max_out = 40,
  mnames = "V",
  seed = 123,
  outlier_type = "x_and_y"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:40,
  distrib_diffs = ombc_lcwm_p1g1_x_and_y$distrib_diffs,
  outlier_num = ombc_lcwm_p1g1_x_and_y$outlier_num
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs <- 
  ombc_lcwm_p1g1_x_and_y$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients

lcwm_p1g1_x_and_y |>
ggplot(aes(
  x = X1, y = Y,
  colour = as.factor(ombc_lcwm_p1g1_x_and_y$labels),
  shape = as.factor(G)
  )) +
  geom_point() +
  geom_abline(
    aes(intercept = coeffs[1], slope = coeffs[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1))
```

### Two-Component, Response Outliers

```{r}
lcwm_p1g2_y_only <- simulate_lcwm(
  n = c(1000, 1000),
  mu = list(c(-1), c(+1)),
  sigma = list(as.matrix(0.2), as.matrix(0.2)),
  beta = list(c(1, 0), c(1, 3)),
  error_sd = c(1, 1),
  outlier_num = c(25, 25),
  outlier_type = "y_only",
  seed = 123,
  crit_val = 0.9999,
  range_multipliers = c(2, 2)
)

ombc_lcwm_p1g2_y_only <- ombc_lcwm(
  xy = lcwm_p1g2_y_only,
  x = lcwm_p1g2_y_only$X1,
  y_formula = Y ~ X1,
  comp_num = 2,
  max_out = 100,
  mnames = "V",
  seed = 123,
  outlier_type = "y_only"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:100,
  distrib_diffs = ombc_lcwm_p1g2_y_only$distrib_diffs,
  outlier_num = ombc_lcwm_p1g2_y_only$outlier_num
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs1 <- 
  ombc_lcwm_p1g2_y_only$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients
coeffs2 <- 
  ombc_lcwm_p1g2_y_only$final_cwm$models[[1]]$GLModel$comp.2$model$coefficients

lcwm_p1g2_y_only |>
  ggplot(aes(
    x = X1, y = Y,
    colour = as.factor(ombc_lcwm_p1g2_y_only$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1, 2)) +
  geom_abline(
    aes(intercept = coeffs1[1], slope = coeffs1[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  geom_abline(
    aes(intercept = coeffs2[1], slope = coeffs2[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 2)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation")
```

### Two-Component, Covariate Outliers

```{r}
lcwm_p1g2_x_only <- simulate_lcwm(
  n = c(1000, 1000),
  mu = list(c(-1), c(+1)),
  sigma = list(as.matrix(0.2), as.matrix(0.2)),
  beta = list(c(1, 0), c(1, 3)),
  error_sd = c(1, 1),
  outlier_num = c(25, 25),
  outlier_type = "x_only",
  seed = 123,
  crit_val = 0.9999,
  range_multipliers = c(2, 2)
)

ombc_lcwm_p1g2_x_only <- ombc_lcwm(
  xy = lcwm_p1g2_x_only,
  x = lcwm_p1g2_x_only$X1,
  y_formula = Y ~ X1,
  comp_num = 2,
  max_out = 100,
  mnames = "V",
  seed = 123,
  outlier_type = "x_only"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:100,
  distrib_diffs = ombc_lcwm_p1g2_x_only$distrib_diffs,
  outlier_num = ombc_lcwm_p1g2_x_only$outlier_num
) |>
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs1 <- 
  ombc_lcwm_p1g2_x_only$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients
coeffs2 <- 
  ombc_lcwm_p1g2_x_only$final_cwm$models[[1]]$GLModel$comp.2$model$coefficients

lcwm_p1g2_x_only |>
  ggplot(aes(
    x = X1, y = Y,
    colour = as.factor(ombc_lcwm_p1g2_x_only$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  geom_abline(
    aes(intercept = coeffs1[1], slope = coeffs1[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  geom_abline(
    aes(intercept = coeffs2[1], slope = coeffs2[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 2)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1, 2))
```

### Two-Component, Combined Outliers

```{r}
lcwm_p1g2_x_and_y <- simulate_lcwm(
  n = c(1000, 1000),
  mu = list(c(-1), c(+1)),
  sigma = list(as.matrix(0.2), as.matrix(0.2)),
  beta = list(c(1, 0), c(1, 3)),
  error_sd = c(1, 1),
  outlier_num = c(25, 25),
  outlier_type = "x_and_y",
  seed = 123,
  crit_val = 0.9999,
  range_multipliers = c(2, 2)
)

ombc_lcwm_p1g2_x_and_y <- ombc_lcwm(
  xy = lcwm_p1g2_x_and_y,
  x = lcwm_p1g2_x_and_y$X1,
  y_formula = Y ~ X1,
  comp_num = 2,
  max_out = 100,
  mnames = "V",
  seed = 123,
  outlier_type = "x_and_y"
)
```

```{r, echo=FALSE, fig.show="hold", out.width="50%"}
data.frame(
  outliers_removed = 0:100,
  distrib_diffs = ombc_lcwm_p1g2_x_and_y$distrib_diffs,
  outlier_num = ombc_lcwm_p1g2_x_and_y$outlier_num
) |> 
  ggplot(aes(x = outliers_removed, y = distrib_diffs)) +
  geom_line() +
  geom_point() +
  geom_vline(aes(xintercept = outlier_num)) +
  labs(x = "Number of Potential Outliers Removed")

coeffs1 <- 
  ombc_lcwm_p1g2_x_and_y$final_cwm$models[[1]]$GLModel$comp.1$model$coefficients
coeffs2 <- 
  ombc_lcwm_p1g2_x_and_y$final_cwm$models[[1]]$GLModel$comp.2$model$coefficients

lcwm_p1g2_x_and_y |>
  ggplot(aes(
    x = X1, y = Y,
    colour = as.factor(ombc_lcwm_p1g2_x_and_y$labels),
    shape = as.factor(G)
  )) +
  geom_point() +
  geom_abline(
    aes(intercept = coeffs1[1], slope = coeffs1[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 1)
  ) +
  geom_abline(
    aes(intercept = coeffs2[1], slope = coeffs2[2]),
    linetype = "solid", colour = ggokabeito::palette_okabe_ito(order = 2)
  ) +
  labs(colour = "outlierMBC", shape = "Simulation") +
  ggokabeito::scale_colour_okabe_ito(order = c(9, 1, 2))
```
